WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-21 articles/2014/Article-WSClean/wsclean.tex
Added Martin's images of Stokes V pulsar detection

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

\subsection{Beam correction}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.

TODO: add Stokes I and V images of polarized source (e.g. Martin's source, if he agrees)

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements significantly. The source code of our new imager is publicly released.
\end{abstract}

@@

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. Each polarization requires a separate run of the $w$-stacking algorithm. Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients, including exoplanets (TODO cite). The pipeline includes WSClean to make the Stokes-parameter images.

Our image-based beam correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam. To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used. The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
